Global patterns of forest autotrophic carbon fluxes

### 2020-03-28 manuscript/manuscript draft.Rmd
rearranged/ re-numbered SI figures
closes #65
woorked on intro

**Carbon cycling in forests worldwide is central to sustaining an estimated #% of Earth's biodiversity (*REF*), while strongly influencing atmospheric CO~2~ and climate** [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity [@waide_relationship_1999] (**REFS**). FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients (**REFS**) [@piao_forest_2010]. Critically, FACF also define forest interactions with Earth's changing climate. Over 69 Gt of CO~2~ cycle through Earth's forests each year [@badgley_terrestrial_2019], and in recent decades their net C sequestration (~2.4 Gt C yr^-1^) offset roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate ([Cavaleri et al. 2015](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.12860); [Rogelj et al. 2018](https://www.ipcc.ch/site/assets/uploads/sites/2/2019/05/SR15_Chapter2_Low_Res.pdf)). However, our ability to draw general macroscopic conclusions regarding global variation in FACF with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables; and typically consider only one or a few variables at a time.

**KAT, work on this:**
**FACF vary with latitude**, showing a general trend of decreasing flux with latitude [@beer_terrestrial_2010; @jung_global_2011]. Studies agree that FACF are lowest in the boreal regions, and increase into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @jung_global_2011]. However, evidence is inconclusive on whether primary productivity continues to increase into the tropics, or whether it plateaus in temperate regions. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity ($GPP$) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019](*[Li & Xiao 2019](https://www.mdpi.com/2072-4292/11/21/2563)*). In contrast, some studies suggest that the highest values of net primary productivity ($NPP$) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], while others find $NPP$ highest in the tropics and decreasing with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity ($ANPP$), finding evidence of a weak negative relationship between $ANPP$ and latitude [@huston_global_2009; @gillman_latitude_2015].

**Climate is a significant driver of FACF across broad spatial scales** [@luyssaert_co_2007; @cleveland_relationships_2011]([Hursh et al. 2017](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.13489)). The majority of studies have focused on exploring the relationships of FACF to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. These variables have the advantage that they describe broad trends in temperature and water availability, and therefore capture a lot of global-scale variation in climate. There is strong evidence that both MAT and MAP show significant positive relationships with FACF [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of FACF across studies. Various measures of primary productivity {**FACF?**} saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity {**FACF?**} are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while GPP increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in GPP saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that NPP increases linearly with temperature. **bbl:maybe remove following sentences** The influence of these climate variables on productivity is further complicated by the possibility of interactive effects occurring between them. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. {**It would be good to add some more citations on soil respiraiton. I'm sure BBL can help.**}

**FACF can be influenced by many other factors as well**, which often act across a range of scales, and may show interactive effects with each other [@cleveland_relationships_2011]. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], biodiversity [@liang_positive_2016], management [@simova_enigma_2017]; nutrient availability [@aragao_above_2009]; and altitude [@girardin_net_2010; @malhi_variation_2017] all impact FACF. On a global scale, we expect that FACF are most strongly influenced by broad climatic gradients. There is evidence that FACFs also respond to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005] in potentially significant ways. Furthermore, MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and FACFs is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between primary productivity and MAT, suggesting that the effect of temperature is due to increased length of growing season, rather than an inherent influence of temperature on FACF.

**The recent development of a global forest carbon database synthesizing multiple variables and including records of stand history (ForC; [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018]) opens up the possibility for a standardized analysis of global scale variation** in multiple FACF and the principle climatic drivers of these patterns. In order to approach this broad topic, we simplify the major gaps in our knowledge to five broad questions and corresponding hypotheses (Table 1). ** see [issue #48 ](https://github.com/forc-db/Global_Productivity/issues/48)** First, we ask how forest autotrophic carbon fluxes (FACF) vary with latitude. We then test how these fluxes relate to MAT and MAP, and additionally how they respond to other, less well-studied, climate variables. Finally, we consider the relationship between FACF and seasonality, considering the role of seasonality in explaining variation in carbon fluxes, and the influence of climate on FACF standardized by growing season length. We use a comprehensive global database of forest carbon fluxes to address the above questions for nine carbon fluxes, allowing for an in-depth exploration of the effect of climate on FACF globally.

@@

@@

We found little evidence that allocation between fluxes varied substantially with latitude or climate (Fig. [**S3**](https://github.com/forc-db/Global_Productivity/issues/65)). Of the 7 FACF ratios regressed against latitude and three climate variables (MAT, MAP, temperature seasonality), there were only four signficant relationships, all with $R^2 \le 0.2$ (Fig. S3). Specifically, the proportion of NPP allocated to $ANPP_{foliage}$ increased weakly with MAT ($R^2=0.20$), and the proportion of $NPP$ allocated aboveground ($ANPP$) decreased weakly with latitude ($R^2=0.11$) and temperature seasonality ($R^2=0.17$), while increasing with MAT ($R^2=0.11$).  There were no significant results from regressing ratios of carbon fluxes against latitude, or against any of the climate variables.

>>>>

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity, while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity [@waide_relationship_1999] (**REFS**). FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients (**REFS**) [@piao_forest_2010]. Critically, FACF also define forest interactions with Earth's changing climate. Over 69 Gt of CO~2~ cycle through Earth's forests each year [@badgley_terrestrial_2019], and in recent decades their net C sequestration (~2.4 Gt C yr^-1^) offset roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate ([Cavaleri et al. 2015](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.12860); [Rogelj et al. 2018](https://www.ipcc.ch/site/assets/uploads/sites/2/2019/05/SR15_Chapter2_Low_Res.pdf)). However, our ability to draw general macroscopic conclusions regarding global variation in FACF with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables; and typically consider only one or a few variables at a time.

FACF decrease with latitude, but it remains unclear whether and how the shape of this relationship varies across fluxes. Studies agree that FACF are lowest in the boreal regions, and increase into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @jung_global_2011]. However, evidence is inconclusive on whether primary productivity continues to increase into the tropics, or whether it plateaus in temperate regions. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity ($GPP$) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019](*[Li & Xiao 2019](https://www.mdpi.com/2072-4292/11/21/2563)*). In contrast, some studies suggest that the highest values of net primary productivity ($NPP$) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], while others find $NPP$ highest in the tropics and decreasing with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity ($ANPP$), finding evidence of a weak negative relationship between $ANPP$ and latitude [@huston_global_2009; @gillman_latitude_2015].

The latitudinal gradient in FACF is primarily driven by climate, which is a significant driver of FACF across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011]([Hursh et al. 2017](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.13489)). The majority of studies have focused on exploring the relationships of FACF to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. While these fail to capture some important aspects of climate such as seasonality, they do describe broad trends in temperature and water availability, and therefore capture a substantial portion of global-scale variation in climate. There is strong evidence that both MAT and MAP show significant positive relationships with FACF [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of FACF across studies. Various measures of primary productivity {**FACF?**} saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity {**FACF?**} are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while $GPP$ increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that $NPP$ increases linearly with temperature. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. {**It would be good to add some more citations on soil respiraiton. I'm sure BBL can help.**}

FACF can be influenced by many other factors as well, which often act across a range of scales, and may show interactive effects with each other [@cleveland_relationships_2011]. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], biodiversity [@liang_positive_2016], management [@simova_enigma_2017], nutrient availability [@aragao_above_2009], and altitude [@girardin_net_2010; @malhi_variation_2017] all impact FACF. On a global scale, we expect that FACF are most strongly influenced by broad climatic gradients. There is evidence that FACFs also respond to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005] in potentially significant ways. Furthermore, MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and FACFs is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between primary productivity and MAT, suggesting that the effect of temperature is due to increased length of growing season, rather than an inherent influence of temperature on FACF.

The recent development of a global forest carbon database synthesizing multiple variables and including records of stand history (ForC; [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018]) opens up the possibility for a standardized analysis of global scale variation in multiple FACF and the principle climatic drivers of these patterns. In order to approach this broad topic, we simplify the major gaps in our knowledge to five broad questions and corresponding hypotheses (Table 1). ** see [issue #48 ](https://github.com/forc-db/Global_Productivity/issues/48)** First, we ask how forest autotrophic carbon fluxes (FACF) vary with latitude. We then test how these fluxes relate to MAT and MAP, and additionally how they respond to other, less well-studied, climate variables. Finally, we consider the relationship between FACF and seasonality, considering the role of seasonality in explaining variation in carbon fluxes, and the influence of climate on FACF standardized by growing season length. We use a comprehensive global database of forest carbon fluxes to address the above questions for nine carbon fluxes, allowing for an in-depth exploration of the effect of climate on FACF globally.

@@

**All analyses were conducted in R (**Version**). Code and results are archived on GitHub....**

@@

We found little evidence that allocation between fluxes varied substantially with latitude or climate (Fig. S3). Of the 7 FACF ratios regressed against latitude and three climate variables (MAT, MAP, temperature seasonality), there were only four signficant relationships, all with $R^2 \le 0.2$ (Fig. S3). Specifically, the proportion of NPP allocated to $ANPP_{foliage}$ increased weakly with MAT ($R^2=0.20$), and the proportion of $NPP$ allocated aboveground ($ANPP$) decreased weakly with latitude ($R^2=0.11$) and temperature seasonality ($R^2=0.17$), while increasing with MAT ($R^2=0.11$).  There were no significant results from regressing ratios of carbon fluxes against latitude, or against any of the climate variables.
